---
title: "YEAR OF BEST POPULAR MOVIES"
author: "Leopold Hillah"
date: "Tuesday, September 23, 2014"
output: html_document
---

---
## ASSUMPTIONS AND ALGORITHM
 
This analysis is about determining the year that the best popular movies were produced.
 
###1. ASSUMPTIONS
 
We define **popular movie** as a movie, which was watched by many people so, received many ratings overall.
 
We define **best movie** as a movie having received a five-star rating on the average (8 or more) or which has received a high proportion of r9 and r10 ratings out of all ratings.
 
###2. MOVIE SELECTION CRITERIA
 
We further define a **success score** as the product of both **popular** rating and **best** rating, standardized on a scale from 1 to 5.
 
We then find movies with the maximum success score and determine their year of production.
 
## DATA PROCESSING
 
###1. RAW DATA
 
The raw data for this project is a tab delimited file available at the following location: [http://had.co.nz/data/movies/](http://had.co.nz/data/movies/)
 
###2. DATA LOADING & PRE-PROCESSING
 
The data load assumes that the tab delimited file is downloaded and stored in the current working directory, whose value can be obtained by the `getwd()` R command. The data is then loaded into the dfmovies data frame using the `read.table` R command.
 
### Setting global options
 
```{r global_config}
# set global chunk options: images will be 24x10 inches
knitr::opts_chunk$set(cache=TRUE, echo=TRUE, message=FALSE, fig.width=24, fig.height=10)
```
 
### Setting up the R environment
 
```{r setting_the_environment}
# Clearing the cache
rm(list = ls())
 
# Loading required libraries
if ((!require(ggplot2)) | (!require(plyr)) | (!require(reshape2))) install.packages('ggplot2', 'plyr', 'reshape2')
 
```
 
### Loading and preprocessing the data
 
```{r data_load_processing}
# Set working directoy here for csv file loading

filepath <- getwd()
 
# Load csv data data frames
dfmovies <- read.table(paste(filepath, "movies.tab", sep="/"), sep="\t", header=TRUE, quote="", comment="", strip.white=TRUE, stringsAsFactors = FALSE)
head(dfmovies)
```
 
Once the data is loaded, it is important to explore the main fields of interest  needed to answer the questions under investigation. Moreover, some initial pre-processing may need to be done first.
 
##### Subsetting data on features of interest #####
 
```{r movie_feature_extraction}
# Extracting features of interest from movie data frame
dfmint <- subset(dfmovies, select=c(year, rating, votes, r1, r2, r3, r4, r5, r6, r7, r8, r9, r10))
```
 
##### Creating movie groups as per above criteria #####

```{r best_movie_list}
# Computing success scores for best movies
 
dfbest <- ddply(dfmint, "year", mutate, best_pop = rating * votes)
```

##### Top best popular movies as per above criteria #####
```{r top_best_movie_list}
# Extract best movie list with rates 8 or more
dftopbest <- arrange(dfbest[dfbest$rating >= 8,], desc(best_pop))
```
 
### Final ranking per year

The final ranking sums up the product of rating and votes aggregated by year, and standardized on a scale from 0 to 100 using the transformation (100 * (value - mean) / (max - min)).

```{r top_best_movie_rank}
# Group data per year and rank them
yeartopbest <- ddply(dftopbest, "year", summarize, best_count = length(year), best_success = sum(best_pop))

# mean and distance from min to max
mu <- mean(yeartopbest$best_success, na.rm = T)
dist <- max(yeartopbest$best_success, na.rm = T) - min(yeartopbest$best_success, na.rm = T)

# Final scaled transformation
yeartopbest <- ddply(yeartopbest, "year", transform, best_success = round(100 * abs(best_success - mu) / dist))
```

### Top 10 years the best popular movies were produced: 1999, 2003, 1994 ...

```{r top_10_best_movie_rank}
# Top 10 years
head(arrange(yeartopbest, desc(best_success)), 10)
```


### Plot showing the years the best popular movies were produced: 1999, 2003, 1994 ...

```{r top_best_movie_graph}
# Group data per year and rank them
p1 <- ggplot(yeartopbest, aes(x=year, y=best_success)) + 
  geom_bar(stat="identity", colour="darkgreen", fill="yellow")+
  xlab("Year")+
  scale_x_continuous(breaks=seq(1900, 2010, 5))+
  scale_y_continuous(breaks=seq(0, 100, 10))+
  ylab("Success Score of Best Popular Movies")+
  theme(text = element_text(size=20),axis.text.x = element_text(angle=30, vjust=1))
 
print(p1)

```
 

