---
title: "Pen-Based Recognition of Handwritten Digits"
author: "Leopold Hillah"
date: "Tuesday, September 30th, 2014"
output: pdf_document
---
 
---
## DATA SOURCE
 
This training data set was made available by the UCI Machine Learning Repository at the folowing location: [http://archive.ics.uci.edu/ml/machine-learning-databases/pendigits/pendigits.tra](http://archive.ics.uci.edu/ml/machine-learning-databases/pendigits/pendigits.tra). The data is stored in csv format.
 
## ANALYSIS
 

###1. DATA LOADING & PRE-PROCESSING
 
The data load assumes that the csv file is downloaded and stored in the current working directory, whose value can be obtained by the `getwd()` R command. The data is then loaded into the dfmovies data frame using the `read.csv` R command.
 
### Setting global options
 
```{r global_config}
# set global chunk options: images will be 24x10 inches
knitr::opts_chunk$set(cache=TRUE, echo=TRUE, message=FALSE, fig.width=24, fig.height=10)
```
 
### Setting up the R environment
 
```{r setting_the_environment}
# Clearing the cache
rm(list = ls())
 
# Loading required libraries
if ((!require(ggplot2)) | (!require(dplyr)) | (!require(reshape2))) install.packages('ggplot2', 'dplyr', 'reshape2')
 
```
 
### Loading and preprocessing the data
 
```{r data_load_processing}
# Set working directoy here for csv file loading
filepath <- getwd()
 
# Load csv data data frames
 
dfpen <- read.csv(paste(filepath, "pendigits.tra", sep="/"), header = FALSE, quote="", comment="", strip.white=TRUE, stringsAsFactors = FALSE)
str(dfpen)
head(dfpen)
```
 
Once the data is loaded, an initial exploration shows that the first 16 variables are input features and the 17th variable is the class. 

**All variables are integer variables.**

### null values per column

We can look at the number of null values in each column of the data set:

```{r null_values}
dfnulls <- colSums(is.na(dfpen))
print(dfnulls)
```

There is no null values in the data set.

### Graph class distribution

We can next look at the distribution of class values among the entire data set using a histogram:
 
```{r class_distribution_graph}
# Graph class distribution
 
dfclass <- summarise (group_by(dfpen, V17), count = n())
print(dfclass)

p1 <- ggplot(dfpen, aes(x=V17)) + 
  geom_histogram(binwidth=0.5, colour="darkblue", fill="white")+
  xlab("Class")+
  ylab("Count")+
  scale_x_continuous(breaks=seq(0, 17, 1))+
  theme(text = element_text(size=20),axis.text.x = element_text(angle=30, vjust=1))
 
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                       
print(p1)
```

**Class value distribution are almost uniform.**

### Correlation among variables

We next look at the correlation among variables (except class or V17):
 
```{r features_correlation}
# Correlation among variables
 
dfcor <- matrix(cor(dfpen[,-17]), ncol=16)
str(dfcor)
# Set diagonals to 0
diag(dfcor) <- 0

for (row in 1:16){
        for (col in 1:16) {
                if (row < col & abs(dfcor[row, col]) > 0.7) print(paste(row, col, dfcor[row, col], sep=":"))
        } 
                } 

```

**Highly correlated variables (70% or more) are :**
-- V4 and V12,
-- V6 and V8,
-- V6 and V14,
-- V8 and V14,
-- V14 and V16
